### 2025-03-13 paper/quarto_ecml/paper.tex
Down to 16 pages without major cutting of content

The trend towards complexity for the sake of performance has come under
serious scrutiny in recent years. At the very cusp of the deep learning
revolution, \cite{szegedy2013intriguing} showed that artificial neural
networks (ANN) are sensitive to adversarial examples: perturbed versions
of data instances that yield vastly different model predictions despite
being ``imperceptible'' in that they are semantically indifferent from
their factual counterparts. Even though some partially effective
mitigation strategies have been proposed---most notably
\textbf{adversarial training} \cite{goodfellow2014explaining}---truly
robust deep learning (DL) remains unattainable even for models that are
considered shallow by today's standards \cite{kolter2023keynote}.

Part of the problem is that the high degrees of freedom provide room for
many solutions that are locally optimal with respect to narrow
objectives \cite{wilson2020case}.\footnote{We follow the standard ML
  convention, where ``degrees of freedom'' refer to the number of
  parameters estimated from data.} Indeed, recent work on the so-called
``lottery ticket hypothesis'' suggests that modern neural networks can
be pruned by up to 90\% while preserving their predictive performance
\cite{frankle2018lottery} and generalizability \cite{morcos2019success}.
Similarly, \cite{zhang2021understanding} showed that state-of-the-art
neural networks are so expressive that they can fit randomly labeled
data. Thus, looking at the predictive performance alone, the solutions
may seem to provide compelling explanations for the data, when in fact
they are based on purely associative, semantically meaningless patterns.
This poses two related challenges. Firstly, there is no dependable way
to verify if such complex representations correspond to meaningful and
plausible explanations. Secondly, even if we could resolve the first
challenge, it remains undecided how to ensure that models can
\emph{only} learn valuable explanations.

@@

\textbf{Specifically, we introduce counterfactual training (CT)}: a
novel training regime explicitly meant to align learned representations
with plausible explanations that comply with user requirements. The
remainder of this paper is structured as follows. Section~\ref{sec-lit}
presents related work, focusing in particular on the link between
adversarial examples and counterfactual explanations. Then follow our
main contributions:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  In Section~\ref{sec-method}, we introduce our methodological framework
  and show theoretically that it can be used to enforce global
  actionability constraints.\\
\item
  Through extensive experiments we demonstrate that CT substantially
  improves explainability without sacrificing predictive performance
  (Section~\ref{sec-experiments}).
\end{enumerate}

We discuss future research and challenges in
Section~\ref{sec-discussion} and conclude in
Section~\ref{sec-conclusion} that CT is a promising new approach towards
making opaque models more trustworthy.

@@

To the best of our knowledge, the proposed framework for counterfactual
training represents the first attempt to use counterfactual explanations
during training to improve model explainability. In high-level terms, we
define model explainability as the extent to which valid explanations
derived for an opaque model are also deemed plausible with respect to
the underlying data and stakeholder requirements; the former means that
the counterfactuals should comply with the distribution of the factual
data, the latter means that they should respect arbitrary (global)
actionability constraints. To make the desiderata for our framework more
concrete, we follow \cite{augustin2020adversarial} in tying the concept
of explainability to the quality of counterfactual explanations that we
can generate for a given model. The authors show that CEs---understood
here as minimal input perturbations that yield some desired model
prediction---are generally more meaningful if the underlying model is
more robust to adversarial examples. We can make intuitive sense of this
finding when looking at adversarial training (AT) through the lens of
representation learning with high degrees of freedom. As argued before,
learned representations may be sensitive to producing implausible
explanations and mispredicting for worst-case counterfactuals (i.e.,
adversarial examples). Thus, by inducing models to ``unlearn''
susceptiblity to such examples, AT can effectively remove implausible
explanations from the solution space.

@@

There have also been several attempts at formalizing the relationship
between counterfactual explanations and adversarial examples (AE).
Pointing to clear similarities in how CEs and AEs are generated,
\cite{freiesleben2022intriguing} makes the case for jointly studying the
opaqueness and robustness problems in representation learning. Formally,
AEs can be seen as the subset of CEs for which misclassification is
achieved \cite{freiesleben2022intriguing}. Similarly,
\cite{pawelczyk2022exploring} show that CEs and AEs are equivalent under
certain conditions and derive theoretical upper bounds on distances
between them.

@@

We say that $\mathbf{M}_\theta$ is \textbf{explainable} to the extent that faithfully generated counterfactuals are plausible and actionable. Formally, we define these properties as follows,

\begin{enumerate}
    \item (Plausibility) $\int^{A} p(\mathbf{x}^\prime|\mathbf{y}^{+})d\mathbf{x} \rightarrow 1$ where $A$ is some small region around $\mathbf{x}^{\prime}$.
    \item (Actionability) Permutations $\Delta$ are subject to some actionability constraints.
    \item (Faithfulness) $\int^{A} p_\theta(\mathbf{x}^\prime|\mathbf{y}^{+})d\mathbf{x} \rightarrow 1$ where $A$ is defined as above.
\end{enumerate}

where $p_\theta(\mathbf{x}|\mathbf{y}^{+})$ denotes the conditional posterior over inputs. 
\end{definition}

@@

\begin{equation}\phantomsection\label{eq-obj}{
\begin{split}
\min_\theta \text{yloss}(\mathbf{M}_\theta(\mathbf{x}),\mathbf{y}) + \lambda_{\text{div}} \text{div}(\mathbf{x}^+,\mathbf{x}_T^\prime,y;\theta) &+ \lambda_{\text{adv}} \text{advloss}(\mathbf{M}_\theta(\mathbf{x}_{t\leq T}^\prime),\mathbf{y}) \\ &+ \lambda_{\text{reg}}\text{ridge}(\mathbf{x}^+,\mathbf{x}_T^\prime,y;\theta)
\end{split}
}\end{equation} where \(\text{yloss}(\cdot)\) is a classification loss
that induces discriminative performance (e.g., cross-entropy). The
second and third terms in Equation~\ref{eq-obj} are explained in detail
below. For now, they can be sufficiently described as inducing
explainability directly and indirectly by penalizing: (1) the
contrastive divergence, \(\text{div}(\cdot)\), between mature
counterfactuals \(\mathbf{x}_T^\prime\) and observed samples
\(\mathbf{x}^+\in\mathcal{X}^+=\{\mathbf{x}:y=y^+\}\) in the target
class \(y^+\), and, (2) the adversarial loss, \(\text{advloss}(.)\),
with respect to nascent counterfactuals \(\mathbf{x}_{t\leq T}^\prime\).
Finally, \(\text{ridge}(\cdot)\) denotes a Ridge penalty
(\(\ell_2\)-norm) that regularizes the magnitude of the energy terms
involved in \(\text{div}(\cdot)\) \cite{du2019implicit}. The trade-off
between the components can be governed by adjusting the strengths of the
penalties \(\lambda_{\text{div}}\), \(\lambda_{\text{adv}}\) and
\(\lambda_{\text{reg}}\).

@@

It is worth highlighting that Prp.\textasciitilde{}\ref{prp-mtblty}
assumes independence of features. This raises a valid concern about the
effect of protecting immutable features in the presence of proxies that
remain unprotected. We address this in Section~\ref{sec-discussion}.

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  To what extent does our proposed counterfactual training objective
  (Equation~\ref{eq-obj}) induce models to learn plausible explanations?
\item
  To what extent does our proposed counterfactual training objective
  (Equation~\ref{eq-obj}) yield more favorable algorithmic recourse
  outcomes in the presence of actionability constraints?
\item
  What are the effects of hyperparameter selection with respect to
  Equation~\ref{eq-obj}?
\end{enumerate}

@@

We observe that for all datasets except \emph{OL} and across all test
settings, the average distance of counterfactuals from observed samples
in the target class is reduced, indicating improved plausibility. The
magnitude of improvements varies by dataset: for the simple synthetic
datasets, distance reductions range from around 20-40\% (\emph{LS},
\emph{Moon}) to almost 60\% (\emph{Circ}). For the real-world tabular
datasets, improvements are generally smaller but still substantial in
many cases with around 10-15\% for \emph{CH}, 11-28\% for \emph{GMSC},
7-8\% for \emph{Cred} and around 3\% for \emph{Adult}. For our only
vision dataset (\emph{MNIST}), distances are reduced by up to 9\%. The
results for our proposed divergence metric are qualitatively similar,
but generally even more pronounced: for the \emph{Circ} dataset,
implausibility is reduced by almost 94\% to virtually zero as we
verified by looking at the absolute outcome. Improvements for other
datasets range from 28\% (\emph{Moon}) to 78\% (\emph{GMSC}). For
\emph{OL} the reduction is negative, consistent with the distance-based
metric. The only dataset, for which our proposed metric disagrees with
the distance-based metric is \emph{MNIST}.

These broad and substantial improvements in plausibility generally do
not come at the cost of decreased predictive performance: test accuracy
for CT is virtually identical to the baseline for \emph{Adult},
\emph{Circ}, \emph{LS}, \emph{Moon} and \emph{OL}, and even slightly
improved for \emph{Cred}. Exceptions to this general pattern are
\emph{MNIST}, \emph{CH} and \emph{GMSC}, for which we observe reduction
in test accuracy of 2, 5 and 15 percentage points, respectively. We note
in this context, that we have not optimized our models for predictive
performance at all and worked with very small networks. In summary, we
find that CT can substantially improve the quality of explanations
learned by models without generally sacrificing predictive accuracy.

@@

We extensively test the impact of three types of hyperparameters on the
proposed training regime. Our complete results are available in the
technical appendix; this section focuses on the main findings.

@@

\textbf{\emph{Other hyperparameters.}} We observe that the effectiveness
and stability of CT is positively associated with the number of
counterfactuals generated during each training epoch. We also confirm
that a higher number of training epochs is beneficial. Interestingly, we
find that it is not necessary to employ CT during the entire training
phase to achieve the desired improvements in explainability. When
training models conventionally during the first 50\% of epochs before
switching to CT for the next 50\% of epochs, we observed positive
results. Put differently, CT may be a way to improve the explainability
of models in a fine-tuning manner.

@@

\textbf{\emph{CT is subject to training instabilities.}} Joint
energy-based models are susceptible to instabilities during training
\cite{grathwohl2020your} and even though we depart from the SGLD-based
sampling, we still encounter major variability in the outcomes. CT is
exposed to two potential sources of instabilities: (1) the energy-based
contrastive divergence term in Equation~\ref{eq-div}, and (2) the
underlying counterfactual explainers. For example,
\cite{altmeyer2024faithful} recognize this to be a challenge for
\emph{ECCCo} and so it may have downstream impacts on our proposed
method. Still, we find that training instabilities can be successfully
mitigated by regularizing energy (\(\lambda_{\text{reg}}\)), generating
a sufficiently large number of counterfactuals during each training
epoch, and including only mature counterfactuals for contrastive
divergence.

\textbf{\emph{CT is sensitive to hyperparameter selection.}} Our method
benefits from tuning certain key hyperparameters (see
Section~\ref{sec-hyperparameters}). In this work, we have relied
exclusively on grid search for this task. Future work on CT could
benefit from investigating more sophisticated approaches towards
hyperparameter tuning. Notably, CT is iterative which makes a variety of
methods applicable, including Bayesian \cite{snoek2012practical} or
gradient-based \cite{franceschi2017forward} optimization.

\subsection{Limitations and Challenges}\label{sec-limitations}

\textbf{\emph{CT increases the training time of models.}} Counterfactual
training promotes explainability through CEs and robustness through AEs
at the cost of longer training times compared to conventional training
regimes. While higher numbers of iterations and counterfactuals per
iteration positively impact the quality of found solutions, they also
increase the required amount of computations. We find that relatively
small grids with 270 settings can take almost four hours for more
demanding datasets on a high-performance computing cluster with 34 2GB
CPUs\footnote{See supplementary appendix for computational details.}.
However, there are three factors that attenuate the impact of this
limitation. First, CT provides counterfactual explanations for the
training samples essentially for free, which may be beneficial in many
ADM systems. Second, we find that CT can retain its value when used as a
``fine-tuning'' training regime for conventionally-trained models.
Third, in principle, CT yields itself to parallel execution, which we
have leveraged for our own experiments.

\textbf{\emph{Immutable features may have proxies.}} We propose an
approach to protect immutable features and thus increase the
actionability of the generated CEs. However, it requires that model
owners define the mutability constraints for (all) features considered
by the model. Even with sufficient domain knowledge to protect all
immutable features, there may exist proxies that are theoretically
mutable (and hence should not be protected) but preserve enough
information about the principals to hinder the protections. As an
example, consider the Adult dataset used in our experiments where the
mutable education status is a proxy for the immutable age, in that the
attainment of degrees is correlated with age. Delineating actionability
is a major undecided challenge in the AR literature
\cite{venkatasubramanian2020philosophical} impacting the capacity of CT
to increase the explainability of the model.

\textbf{\emph{Interventions on features may impact fairness
downstream.}} Related to the point above, we provide a tool that allows
practitioners to modify the sensitivity of a model with respect to
certain features, which may have implication for the fair and equitable
treatment of individuals subject to automated decisions. As protecting a
set of features leads the model to assign higher relative importance to
unprotected features, model owners could misuse our solution by
enforcing explanations based on features that are more difficult to
modify by some (group of) individuals. For example, consider again the
Adult dataset where features such as workclass or education may be more
difficult to change for underpriviledged groups. When applied
irresponsibly, CT could result in an unfairly assigned burden of
recourse \cite{sharma2020certifai}, threatening the equality of
opportunity in the system \cite{bell2024fairness} and potentially
reinforcing social segregation \cite{gao2023impact}. Still, as the
referenced publications indicate, such phenomena are not specific to CT;
all types of ADM solutions without strong external protections have been
recognized to promote harmful power dynamics \cite{maas2023machine}.

@@

State-of-the-art machine learning models are prone to learning complex
representations that cannot be interpreted by humans. Although post-hoc
explainability approaches have attracted major research interest, these
cannot guarantee that the explanations agree with the opaque model's
learned representation of data. As a step towards addressing this
challenge, we introduced counterfactual training, a novel training
regime that incentivizes highly-explainable models. Our approach leads
to explanations that are both plausible---compliant with the underlying
data-generating process---and actionable---compliant with user-specified
mutability constraints---and thus meaningful to their recipients.
Through extensive experiments we demonstrate that CT satisfies its
objectives while preserving the predictive performance of the trained
models. We also find that our approach can be used to fine-tune
conventionally-trained models and achieve similar gains in
explainability. Finally, this work showcases that it is practical to
improve models \emph{and} their explanations at the same time.

>>>>

The trend towards complexity for the sake of performance has come under
serious scrutiny in recent years. At the very cusp of the deep learning
(DL) revolution, \cite{szegedy2013intriguing} showed that artificial
neural networks (ANN) are sensitive to adversarial examples (AEs):
perturbed versions of data instances that yield vastly different model
predictions despite being ``imperceptible'' in that they are
semantically indifferent from their factual counterparts. Even though
some partially effective mitigation strategies have been proposed---most
notably \textbf{adversarial training}
\cite{goodfellow2014explaining}---truly robust deep learning remains
unattainable even for models that are considered ``shallow'' by today's
standards \cite{kolter2023keynote}.

Part of the problem is that the high degrees of freedom provide room for
many solutions that are locally optimal with respect to narrow
objectives \cite{wilson2020case}.\footnote{We follow the standard ML
  convention, where ``degrees of freedom'' refer to the number of
  parameters estimated from data.} Indeed, recent work on the so-called
``lottery ticket hypothesis'' suggests that modern neural networks can
be pruned by up to 90\% while preserving their predictive performance
\cite{frankle2018lottery}. Similarly, \cite{zhang2021understanding}
showed that state-of-the-art neural networks are expressive enough to
fit randomly labeled data. Thus, looking at the predictive performance
alone, the solutions may seem to provide compelling explanations for the
data, when in fact they are based on purely associative, semantically
meaningless patterns. This poses two challenges. Firstly, there is no
dependable way to verify if representations correspond to meaningful,
plausible explanations. Secondly, even if we could resolve the first
challenge, it remains undecided how to ensure that models can
\emph{only} learn valuable explanations.

@@

\textbf{Specifically, we propose counterfactual training (CT)}: a novel
training regime that aligns learned representations with plausible
explanations compliant with user requirements. The remainder of this
paper is structured as follows. Section~\ref{sec-lit} presents related
work, focusing on the link between adversarial examples and
counterfactual explanations. Then follow our main contributions:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  In Section~\ref{sec-method}, we introduce our methodological framework
  and show theoretically that it can be used to enforce global
  actionability constraints.\\
\item
  In Section~\ref{sec-experiments}, through extensive experiments we
  demonstrate that CT substantially improves explainability without
  sacrificing predictive performance.
\end{enumerate}

We discuss the challenges in Section~\ref{sec-discussion} and conclude
in Section~\ref{sec-conclusion} that CT is a promising approach towards
making opaque models more trustworthy.

@@

To the best of our knowledge, the proposed framework for counterfactual
training represents the first attempt to use counterfactual explanations
during training to improve model explainability. In high-level terms, we
define model explainability as the extent to which valid explanations
derived for an opaque model are also deemed plausible with respect to
the underlying data and (global) actionability constraints. To make the
desiderata for our framework more concrete, we follow
\cite{augustin2020adversarial} in tying the concept of explainability to
the quality of CEs that can be generated for a given model. The authors
show that CEs---understood as minimal input perturbations that yield
some desired model prediction---tend to be more meaningful if the
underlying model is more robust to adversarial examples. We can make
intuitive sense of this finding when looking at adversarial training
(AT) through the lens of representation learning with high degrees of
freedom. As argued before, learned representations may be sensitive to
producing implausible explanations and mispredicting for worst-case
counterfactuals (i.e., AEs). Thus, by inducing models to ``unlearn''
susceptiblity to such examples, AT can effectively remove implausible
explanations from the solution space.

@@

There have also been several attempts at formalizing the relationship
between counterfactual explanations and adversarial examples. Pointing
to clear similarities in how CEs and AEs are generated,
\cite{freiesleben2022intriguing} makes the case for jointly studying the
opaqueness and robustness problems in representation learning. Formally,
AEs can be seen as the subset of CEs for which misclassification is
achieved \cite{freiesleben2022intriguing}. Similarly,
\cite{pawelczyk2022exploring} show that CEs and AEs are equivalent under
certain conditions and derive theoretical upper bounds on distances
between them.

@@

We say that $\mathbf{M}_\theta$ is \textbf{explainable} to the extent that faithfully generated counterfactuals are plausible and actionable. We define these properties as follows:

\begin{enumerate}
    \item (Plausibility) $\int^{A} p(\mathbf{x}^\prime|\mathbf{y}^{+})d\mathbf{x} \rightarrow 1$ where $A$ is some small region around $\mathbf{x}^{\prime}$.
    \item (Actionability) Permutations $\Delta$ are subject to some actionability constraints.
    \item (Faithfulness) $\int^{A} p_\theta(\mathbf{x}^\prime|\mathbf{y}^{+})d\mathbf{x} \rightarrow 1$ where $A$ is defined as above.
\end{enumerate}
where $p_\theta(\mathbf{x}|\mathbf{y}^{+})$ denotes the conditional posterior over inputs. 
\end{definition}

@@

\begin{equation}\phantomsection\label{eq-obj}{
\begin{split}
\min_\theta \text{yloss}(\mathbf{M}_\theta(\mathbf{x}),\mathbf{y}) + \lambda_{\text{div}} \text{div}(\mathbf{x}^+,\mathbf{x}_T^\prime,y;\theta) &+ \lambda_{\text{adv}} \text{advloss}(\mathbf{M}_\theta(\mathbf{x}_{t\leq T}^\prime),\mathbf{y}) \\ &+ \lambda_{\text{reg}}\text{ridge}(\mathbf{x}^+,\mathbf{x}_T^\prime,y;\theta)
\end{split}
}\end{equation} where \(\text{yloss}(\cdot)\) is a classification loss
that induces discriminative performance (e.g., cross-entropy). The
second and third terms are explained in detail below. For now, they can
be summarized as inducing explainability directly and indirectly by
penalizing: (1) the contrastive divergence, \(\text{div}(\cdot)\),
between mature counterfactuals \(\mathbf{x}_T^\prime\) and observed
samples \(\mathbf{x}^+\in\mathcal{X}^+=\{\mathbf{x}:y=y^+\}\) in the
target class \(y^+\), and, (2) the adversarial loss,
\(\text{advloss}(.)\), with respect to nascent counterfactuals
\(\mathbf{x}_{t\leq T}^\prime\). Finally, \(\text{ridge}(\cdot)\)
denotes a Ridge penalty (\(\ell_2\)-norm) that regularizes the magnitude
of the energy terms involved in \(\text{div}(\cdot)\)
\cite{du2019implicit}. The trade-off between the components can governed
through penalties \(\lambda_{\text{div}}\), \(\lambda_{\text{adv}}\) and
\(\lambda_{\text{reg}}\).

@@

It is worth highlighting that \mbox{Prp. \ref{prp-mtblty}} assumes
independence of features. This raises a valid concern about the effect
of protecting immutable features in the presence of proxies that remain
unprotected. We address this in Section~\ref{sec-discussion}.

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  To what extent does our proposed counterfactual training objective in
  Equation~\ref{eq-obj} induce models to learn plausible explanations?
\item
  To what extent does our proposed counterfactual training objective in
  Equation~\ref{eq-obj} yield more favorable algorithmic recourse
  outcomes in the presence of actionability constraints?
\item
  What are the effects of hyperparameter selection with respect to
  Equation~\ref{eq-obj}?
\end{enumerate}

@@

We observe that for all datasets except \emph{OL} and across all test
settings, the average distance of counterfactuals from observed samples
in the target class is reduced, indicating improved plausibility. The
magnitude of improvements varies by dataset: for the simple synthetic
datasets, distance reductions range from around 20-40\% (\emph{LS},
\emph{Moon}) to almost 60\% (\emph{Circ}). For the real-world tabular
datasets, improvements are generally smaller but still substantial in
many cases with around 10-15\% for \emph{CH}, 11-28\% for \emph{GMSC},
7-8\% for \emph{Cred} and around 3\% for \emph{Adult}. For our only
vision dataset (\emph{MNIST}), distances are reduced by up to 9\%. The
results for our proposed divergence metric are qualitatively similar,
but generally even more pronounced: for the \emph{Circ} dataset,
implausibility is reduced by almost 94\% to virtually zero as we
verified by looking at the absolute outcome. Improvements for other
datasets range from 28\% (\emph{Moon}) to 78\% (\emph{GMSC}). For
\emph{OL} the reduction is negative, consistent with the distance-based
metric. \emph{MNIST} is the only dataset for which the two metrics
disagree.

These broad and substantial improvements in plausibility generally do
not come at the cost of decreased predictive performance: test accuracy
for CT is virtually identical to the baseline for \emph{Adult},
\emph{Circ}, \emph{LS}, \emph{Moon} and \emph{OL}, and even slightly
improved for \emph{Cred}. Exceptions to this general pattern are
\emph{MNIST}, \emph{CH} and \emph{GMSC}, for which we observe reduction
in test accuracy of 2, 5 and 15 percentage points, respectively. We note
in this context, that we have not optimized our models for predictive
performance at all and worked with very small networks. In summary, we
find that CT can substantially improve the quality of explanations
learned by models without sacrificing predictive accuracy.

@@

We test in-depth the impact of three types of hyperparameters; our
complete results are in the appendix.

@@

\textbf{\emph{Other hyperparameters.}} We observe that the effectiveness
and stability of CT is positively associated with the number of
counterfactuals generated during each training epoch. We also confirm
that a higher number of training epochs is beneficial. Interestingly, we
observed desired improvements in explainability when CT was combined
with conventional training and applied only for the final 50\% of epochs
of the complete training process. Put differently, CT may be a way to
improve the explainability of models in a fine-tuning manner.

@@

\textbf{\emph{CT is subject to training instabilities.}} JEMs are
susceptible to instabilities during training \cite{grathwohl2020your}
and even though we depart from the SGLD-based sampling, we still
encounter major variability in the outcomes. CT is exposed to two
potential sources of instabilities: (1) the energy-based contrastive
divergence term in Equation~\ref{eq-div}, and (2) the underlying
counterfactual explainers. Still, we find that training instabilities
can be successfully mitigated by regularizing energy
(\(\lambda_{\text{reg}}\)), generating sufficiently many counterfactuals
during each training epoch, and including only mature counterfactuals
for contrastive divergence.

\textbf{\emph{CT is sensitive to hyperparameter selection.}} Our method
benefits from the tuning of certain key hyperparameters (see
Section~\ref{sec-hyperparameters}). In this work, we have relied
exclusively on grid search for this task. Future work on CT could
benefit from investigating more sophisticated approaches towards
hyperparameter tuning. Notably, CT is iterative which makes a variety of
methods applicable, including Bayesian \cite{snoek2012practical} or
gradient-based \cite{franceschi2017forward} optimization.

\subsection{Current Limitations}\label{sec-limitations}

\textbf{\emph{CT increases the training time of models.}} CT can be more
time-consuming than conventional training regimes. While higher numbers
of CEs per iteration positively impact the quality of solutions, they
also increase the amount of computations. Relatively small grids with
270 settings can take almost four hours for more demanding datasets on a
high-performance computing cluster with 34 2GB CPUs.\footnote{See
  supplementary appendix for computational details.} Three factors
attenuate this effect. First, CT amortizes the cost of CEs for the
training samples. Second, we find that it can retain its value when used
as a ``fine-tuning'' technique for conventionally-trained models. Third,
it yields itself to parallel execution, which we have leveraged for our
own experiments.

\textbf{\emph{Immutable features may have proxies.}} We propose an
approach to protect immutable features and thus increase the
actionability of the generated CEs. However, it requires that model
owners define the mutability constraints for (all) features considered
by the model. Even if all immutable features are protected, there may
exist proxies that are mutable (and hence should not be protected) but
preserve enough information about the principals to hinder the
protections. Delineating actionability is a major undecided challenge in
the AR literature \cite{venkatasubramanian2020philosophical} impacting
the capacity of CT to increase the explainability of the model.

\textbf{\emph{Interventions on features may impact fairness.}} We
provide a tool that allows practitioners to modify the sensitivity of a
model with respect to certain features, which may have implication for
the fair and equitable treatment of decision subjects. As protecting a
set of features leads the model to assign higher relative importance to
unprotected features, model owners could misuse our solution by
enforcing explanations based on features that are more difficult to
modify by some (group of) individuals. For example, consider the Adult
dataset used in our experiments, where \emph{workclass} or
\emph{education} may be more difficult to change for underpriviledged
groups. When applied irresponsibly, CT could result in an unfairly
assigned burden of recourse \cite{sharma2020certifai}, threatening the
equality of opportunity in the system \cite{bell2024fairness}. Still,
these phenomena are not specific to CT.

@@

State-of-the-art machine learning models are prone to learning complex
representations that cannot be interpreted by humans and existing
post-hoc explainability approaches cannot guarantee that the
explanations agree with the model's learned representation of data. As a
step towards addressing this challenge, we introduced counterfactual
training, a novel training regime that incentivizes highly-explainable
models. Our approach leads to explanations that are both
plausible---compliant with the underlying data-generating process---and
actionable---compliant with user-specified mutability constraints---and
thus meaningful to their recipients. Through extensive experiments we
demonstrate that CT satisfies its objectives while preserving the
predictive performance of the models. Our approach can also be used to
fine-tune conventionally-trained models and achieve similar gains in
explainability. Finally, this work showcases that it is practical to
improve models \emph{and} their explanations at the same time.

### 2025-03-13 paper/sections/conclusion.qmd
Down to 16 pages without major cutting of content

# Conclusion {#sec-conclusion}
State-of-the-art machine learning models are prone to learning complex representations that cannot be interpreted by humans. Although post-hoc explainability approaches have attracted major research interest, these cannot guarantee that the explanations agree with the opaque model's learned representation of data. As a step towards addressing this challenge, we introduced counterfactual training, a novel training regime that incentivizes highly-explainable models. Our approach leads to explanations that are both plausible---compliant with the underlying data-generating process---and actionable---compliant with user-specified mutability constraints---and thus meaningful to their recipients. Through extensive experiments we demonstrate that CT satisfies its objectives while preserving the predictive performance of the trained models. We also find that our approach can be used to fine-tune conventionally-trained models and achieve similar gains in explainability. Finally, this work showcases that it is practical to improve models *and* their explanations at the same time.

>>>>

# Conclusion {#sec-conclusion}
State-of-the-art machine learning models are prone to learning complex representations that cannot be interpreted by humans and existing post-hoc explainability approaches cannot guarantee that the explanations agree with the model's learned representation of data. As a step towards addressing this challenge, we introduced counterfactual training, a novel training regime that incentivizes highly-explainable models. Our approach leads to explanations that are both plausible---compliant with the underlying data-generating process---and actionable---compliant with user-specified mutability constraints---and thus meaningful to their recipients. Through extensive experiments we demonstrate that CT satisfies its objectives while preserving the predictive performance of the models. Our approach can also be used to fine-tune conventionally-trained models and achieve similar gains in explainability. Finally, this work showcases that it is practical to improve models *and* their explanations at the same time.

### 2025-03-13 paper/sections/discussion.qmd
Fix references
Down to 16 pages without major cutting of content

***CT is subject to training instabilities.*** Joint energy-based models are susceptible to instabilities during training [@grathwohl2020your] and even though we depart from the SGLD-based sampling, we still encounter major variability in the outcomes. CT is exposed to two potential sources of instabilities: (1) the energy-based contrastive divergence term in @eq-div, and (2) the underlying counterfactual explainers. For example, @altmeyer2023faithful recognize this to be a challenge for *ECCCo* and so it may have downstream impacts on our proposed method. Still, we find that training instabilities can be successfully mitigated by regularizing energy ($\lambda_{\text{reg}}$), generating a sufficiently large number of counterfactuals during each training epoch, and including only mature counterfactuals for contrastive divergence.

***CT is sensitive to hyperparameter selection.*** Our method benefits from tuning certain key hyperparameters (see @sec-hyperparameters). In this work, we have relied exclusively on grid search for this task. Future work on CT could benefit from investigating more sophisticated approaches towards hyperparameter tuning. Notably, CT is iterative which makes a variety of methods applicable, including Bayesian [e.g., @snoek2012practical] or gradient-based [e.g., @franceschi2017forward] optimization.

## Limitations and Challenges {#sec-limitations}

***CT increases the training time of models.*** Counterfactual training promotes explainability through CEs and robustness through AEs at the cost of longer training times compared to conventional training regimes. While higher numbers of iterations and counterfactuals per iteration positively impact the quality of found solutions, they also increase the required amount of computations. We find that relatively small grids with 270 settings can take almost four hours for more demanding datasets on a high-performance computing cluster with 34 2GB CPUs^[See supplementary appendix for computational details.]. However, there are three factors that attenuate the impact of this limitation. First, CT provides counterfactual explanations for the training samples essentially for free, which may be beneficial in many ADM systems. Second, we find that CT can retain its value when used as a "fine-tuning" training regime for conventionally-trained models. Third, in principle, CT yields itself to parallel execution, which we have leveraged for our own experiments.

***Immutable features may have proxies.*** We propose an approach to protect immutable features and thus increase the actionability of the generated CEs. However, it requires that model owners define the mutability constraints for (all) features considered by the model. Even with sufficient domain knowledge to protect all immutable features, there may exist proxies that are theoretically mutable (and hence should not be protected) but preserve enough information about the principals to hinder the protections. As an example, consider the Adult dataset used in our experiments where the mutable education status is a proxy for the immutable age, in that the attainment of degrees is correlated with age. Delineating actionability is a major undecided challenge in the AR literature [see, e.g., @venkatasubramanian2020philosophical] impacting the capacity of CT to increase the explainability of the model.

***Interventions on features may impact fairness downstream.***  Related to the point above, we provide a tool that allows practitioners to modify the sensitivity of a model with respect to certain features, which may have implication for the fair and equitable treatment of individuals subject to automated decisions. As protecting a set of features leads the model to assign higher relative importance to unprotected features, model owners could misuse our solution by enforcing explanations based on features that are more difficult to modify by some (group of) individuals. For example, consider again the Adult dataset where features such as workclass or education may be more difficult to change for underpriviledged groups. When applied irresponsibly, CT could result in an unfairly assigned burden of recourse [e.g., @sharma2020certifai], threatening the equality of opportunity in the system [@bell2024fairness] and potentially reinforcing social segregation [@gao2023impact]. Still, as the referenced publications indicate, such phenomena are not specific to CT; all types of ADM solutions without strong external protections have been recognized to promote harmful power dynamics [@maas2023machine].

>>>>

***CT is subject to training instabilities.*** JEMs are susceptible to instabilities during training [@grathwohl2020your] and even though we depart from the SGLD-based sampling, we still encounter major variability in the outcomes. CT is exposed to two potential sources of instabilities: (1) the energy-based contrastive divergence term in @eq-div, and (2) the underlying counterfactual explainers. Still, we find that training instabilities can be successfully mitigated by regularizing energy ($\lambda_{\text{reg}}$), generating sufficiently many counterfactuals during each training epoch, and including only mature counterfactuals for contrastive divergence.

***CT is sensitive to hyperparameter selection.*** Our method benefits from the tuning of certain key hyperparameters (see @sec-hyperparameters). In this work, we have relied exclusively on grid search for this task. Future work on CT could benefit from investigating more sophisticated approaches towards hyperparameter tuning. Notably, CT is iterative which makes a variety of methods applicable, including Bayesian [e.g., @snoek2012practical] or gradient-based [e.g., @franceschi2017forward] optimization.

## Current Limitations {#sec-limitations}

***CT increases the training time of models.*** CT can be more time-consuming than conventional training regimes. While higher numbers of CEs per iteration positively impact the quality of solutions, they also increase the amount of computations. Relatively small grids with 270 settings can take almost four hours for more demanding datasets on a high-performance computing cluster with 34 2GB CPUs.^[See supplementary appendix for computational details.] Three factors attenuate this effect. First, CT amortizes the cost of CEs for the training samples. Second, we find that it can retain its value when used as a "fine-tuning" technique for conventionally-trained models. Third, it yields itself to parallel execution, which we have leveraged for our own experiments.

***Immutable features may have proxies.*** We propose an approach to protect immutable features and thus increase the actionability of the generated CEs. However, it requires that model owners define the mutability constraints for (all) features considered by the model. Even if all immutable features are protected, there may exist proxies that are mutable (and hence should not be protected) but preserve enough information about the principals to hinder the protections. Delineating actionability is a major undecided challenge in the AR literature [see, e.g., @venkatasubramanian2020philosophical] impacting the capacity of CT to increase the explainability of the model.

***Interventions on features may impact fairness.***  We provide a tool that allows practitioners to modify the sensitivity of a model with respect to certain features, which may have implication for the fair and equitable treatment of decision subjects. As protecting a set of features leads the model to assign higher relative importance to unprotected features, model owners could misuse our solution by enforcing explanations based on features that are more difficult to modify by some (group of) individuals. For example, consider the Adult dataset used in our experiments, where *workclass* or *education* may be more difficult to change for underpriviledged groups. When applied irresponsibly, CT could result in an unfairly assigned burden of recourse [e.g., @sharma2020certifai], threatening the equality of opportunity in the system [@bell2024fairness]. Still, these phenomena are not specific to CT.
